---
title: "Fixed or random?"
author: "Anna Quaglieri"
date: "13/12/2017"
output:
  github_document:
    toc: true
  html_document:
    toc: true
    theme: readable
  ioslides_presentation: default
  slidy_presentation: default
---


```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(lme4)
library(ggplot2)
```

# Confounded design 

```{r}
mu = 0
d = 8
b = 4
```


1. $E(x_i) = \mu$
2. $E(y_i) = \mu + d$
3. $E(x_i') = \mu + b$

```{r}
set.seed(100)
x <- rnorm(25,mean = mu,sd = 1)
set.seed(100)
y <- rnorm(25,mean = mu + d,sd = 1.7)
set.seed(100)
x_1 <- rnorm(25,mean = mu + b,sd = 1.4)

value <- c(x,y,x_1)
x <- data.frame(value = value,
  Treatment = c(rep(c(0,1,0),times=c(25,25,25))),
                blocks = rep(c(1,2),times=c(50,25)))

ggplot(x,aes(x=factor(blocks),y=value,fill=factor(Treatment))) + geom_boxplot() 
```

## Block as a known fixed effect

```{r}
summary(lm(value ~ factor(Treatment) + factor(blocks),data=x))
```

* $\mu$

```{r}
mean(x$value[x$Treatment==0 & x$blocks == 1])
```

* *disease effect*

$d=\bar{y} - \bar{x}$

```{r}
mean(x$value[x$Treatment==1 & x$blocks == 1]) - mean(x$value[x$Treatment==0 & x$blocks == 1])
```

* *batch effect* $b=\bar{x}' - \bar{x}$



```{r}
mean(x$value[x$Treatment==0 & x$blocks == 2]) - mean(x$value[x$Treatment==0 & x$blocks == 1])
```

## What if we removed block 2?

```{r}
summary(lm(value ~ factor(Treatment),data=x[x$blocks == 1,]))
```

* The second batch does not provide any information in the estimate of the disease effect.
* Is it still worth putting it in?

## The estimate of the variance of the effect size depends on the residuals

* Estimate of the residual variance after fitting the model. 

${\sigma}^2 = \frac{sum(residuals^2)}{n - estimated \; betas}$

```{r}
mod1 <- lm(value ~ factor(Treatment) + factor(blocks),data=x)

sigma <- sum(mod1$residuals^2)/(nrow(x)-3)
x2 <- x
x2[,2] <- as.numeric(as.character(x2[,2]))
x2[,3] <- as.numeric(as.character(x2[,3]))
X <- as.matrix(cbind(1,x2[,2:3]))

unscaled_var <- solve(t(X) %*% X)

sqrt(unscaled_var*sigma)

summary(mod1)
```

* If we only use block 1

```{r}
mod1 <- lm(value ~ factor(Treatment),data=x[x$blocks == 1,])
sigma <- sum(mod1$residuals^2)/(nrow(x[x$blocks == 0,])-2)
x2 <- x[x$blocks == 0,]
x2[,2] <- as.numeric(as.character(x2[,2]))
X <- as.matrix(cbind(1,x2[,2]))

unscaled_var <- solve(t(X) %*% X)

sqrt(unscaled_var*sigma)

summary(mod1)
```

* Which is equivalent to estimating:

$V(disease \; effect) = V(mean \; in \; disease) -  V(mean \; in \; controls)$

$V(mean \; in \; disease) = \frac{{\sigma\; disease}^2}{n}$

```{r}
nc <- length(x$value[x$Treatment==0 & x$blocks == 0])
vc <- var(x$value[x$Treatment==0 & x$blocks == 0])/nc

nd <- length(x$value[x$Treatment==1 & x$blocks == 0])
vd <- var(x$value[x$Treatment==1 & x$blocks == 0])/nd

sqrt(vd + vc)
```

* What if we don't include *blocks*?

```{r}
mod1 <- lm(value ~ factor(Treatment),data=x)

summary(mod1)
```

## Block as random effect

```{r}
x$blocks <- factor(x$blocks)
fm1 <- lmer(value ~ factor(Treatment) + (1 | blocks), data = x)
summary(fm1)
```




# Compare fixed random models

* Only group one contains both treatments

```{r}
library(lme4)
library(ggplot2)
```

## Distribution of b_j at changing sigma2_b

```{r}
set.seed(100)
plot(density(rnorm(1000,mean = 0,sd = sqrt(4))),xlab="b_j",ylab="density",main="1k Random sampling from b_j distribution",xlim=c(-20,20))
lines(density(rnorm(1000,mean = 0,sd = sqrt(40))),col=2)
lines(density(rnorm(1000,mean = 0,sd = sqrt(400))),col=4)
```

## sigma2_b = sigma2_e

```{r}
mu <- 0 # mean in controls in batch1
d <- 8 # disease effect
sigma2_b <- 4 
sigma2_e <- 4
```

* Random $b_{js}$. The random intercepts below are quite close to each other, meaning that the batch effect might not be that small. The random effects could be increases by increasing $sigma_b$

```{r}
set.seed(100)
b <- rnorm(2,mean = 0,sd = sqrt(sigma2_b))

b[1]
b[2]
```

```{r}
set.seed(100)
x <- rnorm(25,mean = mu + b[1],sd = sqrt(sigma2_e))
set.seed(100)
y <- rnorm(25,mean = mu + d + b[1],sd = sqrt(sigma2_e))
set.seed(3546)
x_1 <- rnorm(25,mean = mu + b[2],sd = sqrt(sigma2_e))

value <- c(x,y,x_1)
datax <- data.frame(value = value,
  Treatment = c(rep(c(0,1,0),times=c(25,25,25))),
                blocks = rep(c(1,2),times=c(50,25)))

ggplot(datax,aes(x=factor(blocks),y=value,fill=factor(Treatment))) + geom_boxplot()
```

```{r}
datax$Treatment <- factor(datax$Treatment) 
datax$blocks <- factor(datax$blocks) 
table(datax$Treatment,datax$blocks)
```


```{r}
summary(lm(value ~ Treatment + blocks,data=datax))
```

$\bar{y} - \bar{x}$

```{r}
mean(datax$value[datax$Treatment == 1 & datax$blocks == 1]) - mean(datax$value[datax$Treatment == 0 & datax$blocks == 1])
```


```{r}
datax$blocks <- factor(datax$blocks)
fm1 <- lmer(value ~ factor(Treatment) + (1 | blocks), data = datax)
summary(fm1)
```


## sigma2_b = 10*sigma2_e

```{r}
mu <- 0 # mean in controls in batch1
d <- 8 # disease effect
sigma2_e <- 4
sigma2_b <- 10*sigma2_e
```


```{r}
set.seed(100)
b <- rnorm(2,mean = 0,sd = sqrt(sigma2_b))

b[1]
b[2]
```

```{r}
set.seed(100)
x <- rnorm(25,mean = mu + b[1],sd = sqrt(sigma2_e))
set.seed(100)
y <- rnorm(25,mean = mu + d + b[1],sd = sqrt(sigma2_e))
set.seed(3546)
x_1 <- rnorm(25,mean = mu + b[2],sd = sqrt(sigma2_e))

value <- c(x,y,x_1)
datax <- data.frame(value = value,
  Treatment = c(rep(c(0,1,0),times=c(25,25,25))),
                blocks = rep(c(1,2),times=c(50,25)))

ggplot(datax,aes(x=factor(blocks),y=value,fill=factor(Treatment))) + geom_boxplot()
```

```{r}
datax$Treatment <- factor(datax$Treatment) 
datax$blocks <- factor(datax$blocks) 
table(datax$Treatment,datax$blocks)
```


```{r}
summary(lm(value ~ Treatment + blocks,data=datax))
```


```{r}
datax$blocks <- factor(datax$blocks)
fm1 <- lmer(value ~ factor(Treatment) + (1 | blocks), data = datax)
summary(fm1)
```


## sigma2_b = 100*sigma2_e

```{r}
mu <- 0 # mean in controls in batch1
d <- 8 # disease effect
sigma2_e <- 4 
sigma2_b <- 100*sigma2_e
```

* Distribution of batch effects

```{r}
set.seed(100)
b <- rnorm(2,mean = 0,sd = sqrt(sigma2_b))

b[1]
b[2]
```

```{r}
set.seed(100)
x <- rnorm(25,mean = mu + b[1],sd = sqrt(sigma2_e))
set.seed(100)
y <- rnorm(25,mean = mu + d + b[1],sd = sqrt(sigma2_e))
set.seed(3546)
x_1 <- rnorm(25,mean = mu + b[2],sd = sqrt(sigma2_e))

value <- c(x,y,x_1)
datax <- data.frame(value = value,
  Treatment = c(rep(c(0,1,0),times=c(25,25,25))),
                blocks = rep(c(1,2),times=c(50,25)))

ggplot(datax,aes(x=factor(blocks),y=value,fill=factor(Treatment))) + geom_boxplot()
```

```{r}
datax$Treatment <- factor(datax$Treatment) 
datax$blocks <- factor(datax$blocks) 
table(datax$Treatment,datax$blocks)
```


```{r}
summary(lm(value ~ Treatment + blocks,data=datax))
```


```{r}
datax$blocks <- factor(datax$blocks)
fm1 <- lmer(value ~ factor(Treatment) + (1 | blocks), data = datax)
summary(fm1)
```


## Comments

* In the examples above we increased $sigma^2_b$ to allow for higher variability between batches. It is interesting to observe that, as the variability between batches tends to infinity, the effect size estimated by the mixed model approaches the fixed effect estimate. 

* One intuitive way to think about it 



# Simulate four batches

## High disease effect and low between batches variability

```{r}
mu <- 0 # mean in controls in batch1
d <- 30 # disease effect
sigma2_e <- 2 
sigma2_b <- 2

set.seed(100)
b <- rnorm(5,mean = 0,sd = sqrt(sigma2_b))

b[1]
b[2]
b[3]
b[4]
b[5]
```


```{r}
set.seed(100)
x <- rnorm(25,mean = mu + b[1],sd = sqrt(sigma2_e))
set.seed(100)
y <- rnorm(25,mean = mu + d + b[1],sd = sqrt(sigma2_e))
set.seed(3546)
x_1 <- rnorm(25,mean = mu + b[2],sd = sqrt(sigma2_e))
set.seed(100)
x_2 <- rnorm(25,mean = mu + b[3],sd = sqrt(sigma2_e))
set.seed(100)
y_2 <- rnorm(25,mean = mu + d + b[3],sd = sqrt(sigma2_e))
set.seed(3546)
x_3 <- rnorm(25,mean = mu + b[4],sd = sqrt(sigma2_e))
set.seed(100)
y_3 <- rnorm(25,mean = mu + d + b[4],sd = sqrt(sigma2_e))
set.seed(100)
x_4 <- rnorm(25,mean = mu + b[5],sd = sqrt(sigma2_e))

value <- c(x,y,x_1,x_2,y_2,x_3,y_3,x_4)
datax <- data.frame(value = value,
  Treatment = c(rep(c(0,1,0,0,1,0,1,0),times=rep(25,8))),
                blocks = rep(c(1,2,3,4,5),times=c(50,25,50,50,25)))

ggplot(datax,aes(x=factor(blocks),y=value,fill=factor(Treatment))) + geom_boxplot()
```


## High disease effect and high between batches variability

```{r}
mu <- 0 # mean in controls in batch1
d <- 30 # disease effect
sigma2_e <- 2 
sigma2_b <- 100*sigma2_e

set.seed(100)
b <- rnorm(5,mean = 0,sd = sqrt(sigma2_b))

b[1]
b[2]
b[3]
b[4]
b[5]
```


```{r}
set.seed(100)
x <- rnorm(25,mean = mu + b[1],sd = sqrt(sigma2_e))
set.seed(100)
y <- rnorm(25,mean = mu + d + b[1],sd = sqrt(sigma2_e))
set.seed(3546)
x_1 <- rnorm(25,mean = mu + b[2],sd = sqrt(sigma2_e))
set.seed(100)
x_2 <- rnorm(25,mean = mu + b[3],sd = sqrt(sigma2_e))
set.seed(100)
y_2 <- rnorm(25,mean = mu + d + b[3],sd = sqrt(sigma2_e))
set.seed(3546)
x_3 <- rnorm(25,mean = mu + b[4],sd = sqrt(sigma2_e))
set.seed(100)
y_3 <- rnorm(25,mean = mu + d + b[4],sd = sqrt(sigma2_e))
set.seed(100)
x_4 <- rnorm(25,mean = mu + b[5],sd = sqrt(sigma2_e))

value <- c(x,y,x_1,x_2,y_2,x_3,y_3,x_4)
datax <- data.frame(value = value,
  Treatment = c(rep(c(0,1,0,0,1,0,1,0),times=rep(25,8))),
                blocks = rep(c(1,2,3,4,5),times=c(50,25,50,50,25)))

ggplot(datax,aes(x=factor(blocks),y=value,fill=factor(Treatment))) + geom_boxplot()
```

```{r}
summary(lm(value ~ Treatment + blocks,data=datax))
```

* Resisuals after using fixed effects

```{r}
mod <- lm(value ~ blocks,data=datax)
datax$res <- mod$residuals
ggplot(datax,aes(x=factor(blocks),y=res,fill=factor(Treatment))) + geom_boxplot()
```


```{r}
fm1 <- lmer(value ~ factor(Treatment) + (1 | blocks), data = datax)
summary(fm1)
```


## Low disease effect and high between batches variability

```{r}
mu <- 0 # mean in controls in batch1
d <- 4 # disease effect
sigma2_e <- 2 
sigma2_b <- 100*sigma2_e

set.seed(100)
b <- rnorm(5,mean = 0,sd = sqrt(sigma2_b))

b[1]
b[2]
b[3]
b[4]
b[5]
```


```{r}
set.seed(100)
x <- rnorm(25,mean = mu + b[1],sd = sqrt(sigma2_e))
set.seed(100)
y <- rnorm(25,mean = mu + d + b[1],sd = sqrt(sigma2_e))
set.seed(3546)
x_1 <- rnorm(25,mean = mu + b[2],sd = sqrt(sigma2_e))
set.seed(100)
x_2 <- rnorm(25,mean = mu + b[3],sd = sqrt(sigma2_e))
set.seed(100)
y_2 <- rnorm(25,mean = mu + d + b[3],sd = sqrt(sigma2_e))
set.seed(3546)
x_3 <- rnorm(25,mean = mu + b[4],sd = sqrt(sigma2_e))
set.seed(100)
y_3 <- rnorm(25,mean = mu + d + b[4],sd = sqrt(sigma2_e))
set.seed(100)
x_4 <- rnorm(25,mean = mu + b[5],sd = sqrt(sigma2_e))

value <- c(x,y,x_1,x_2,y_2,x_3,y_3,x_4)
datax <- data.frame(value = value,
  Treatment = c(rep(c(0,1,0,0,1,0,1,0),times=rep(25,8))),
                blocks = rep(c(1,2,3,4,5),times=c(50,25,50,50,25)))

ggplot(datax,aes(x=factor(blocks),y=value,fill=factor(Treatment))) + geom_boxplot()
```

```{r}
summary(lm(value ~ Treatment + blocks,data=datax))
```

```{r}
fm1 <- lmer(value ~ factor(Treatment) + (1 | blocks), data = datax)
summary(fm1)
```


## Low disease effect and low between batches variability

```{r}
mu <- 0 # mean in controls in batch1
d <- 4 # disease effect
sigma2_e <- 2 
sigma2_b <- sigma2_e

set.seed(100)
b <- rnorm(5,mean = 0,sd = sqrt(sigma2_b))

b[1]
b[2]
b[3]
b[4]
b[5]
```


```{r}
set.seed(100)
x <- rnorm(25,mean = mu + b[1],sd = sqrt(sigma2_e))
set.seed(100)
y <- rnorm(25,mean = mu + d + b[1],sd = sqrt(sigma2_e))
set.seed(3546)
x_1 <- rnorm(25,mean = mu + b[2],sd = sqrt(sigma2_e))
set.seed(100)
x_2 <- rnorm(25,mean = mu + b[3],sd = sqrt(sigma2_e))
set.seed(100)
y_2 <- rnorm(25,mean = mu + d + b[3],sd = sqrt(sigma2_e))
set.seed(3546)
x_3 <- rnorm(25,mean = mu + b[4],sd = sqrt(sigma2_e))
set.seed(100)
y_3 <- rnorm(25,mean = mu + d + b[4],sd = sqrt(sigma2_e))
set.seed(100)
x_4 <- rnorm(25,mean = mu + b[5],sd = sqrt(sigma2_e))

value <- c(x,y,x_1,x_2,y_2,x_3,y_3,x_4)
datax <- data.frame(value = value,
  Treatment = c(rep(c(0,1,0,0,1,0,1,0),times=rep(25,8))),
                blocks = rep(c(1,2,3,4,5),times=c(50,25,50,50,25)))

ggplot(datax,aes(x=factor(blocks),y=value,fill=factor(Treatment))) + geom_boxplot()
```

```{r}
summary(lm(value ~ Treatment + blocks,data=datax))
```

```{r}
fm1 <- lmer(value ~ factor(Treatment) + (1 | blocks), data = datax)
summary(fm1)
```
